Impact of Federal Party Leadership Changes on Donation Patterns

### 2024-12-03 paper/paper.qmd
update cleaning code

A key question in political finance is whether a party's position in power affects the amount of financial support it receives. Studies in other countries, such as the  @fouirnaies2014financial conducted in the United States, have shown that incumbents often benefit from more donations, particularly from interest groups seeking to maintain influence. However, there has been less focus on whether this holds true in Canada, where the political system and donation regulations differ significantly. This paper aims to fill that gap by analyzing how a party's power status influences donor behaviour in federal and Ontario provincial elections.

@@

## Overview

The dataset used in this analysis comprises political donation records
spanning from 2013 to 2024, covering both the federal level and the
province of Ontario in Canada. These records were sourced from the
Investigative Journalism Foundation (IJF), which systematically collects
and maintains comprehensive data on political donations across various
Canadian jurisdictions. The IJF gathers daily updates by monitoring
elections agency websites at the federal level and within each province
and territory, ensuring the dataset remains current and robust. Although
the IJF's repository includes historical data dating back to 1993, this
study focuses exclusively on the 2013–2024 period to capture recent
trends and changes in political financing dynamics (Investigative
Journalism Foundation 2024).

The analysis was conducted using the statistical programming language R
(R Core Team 2024) and several packages for data manipulation and
visualization, including `dplyr` (Wickham et al. 2023), `ggplot2`
(Wickham 2016), `readr` (Wickham et al. 2023), and `tidyr` (Wickham et
al. 2023). \[UPDATE REFERENCES\]

@@

The dataset used in this analysis originates from the Investigative
Journalism Foundation (IJF), which compiles political donation records
from Canadian federal elections and the province of Ontario, covering
the period from 2013 to 2024. The IJF collects this data by monitoring
official elections agency websites at both the federal level and within
each province and territory. Donations initially reported in various
formats, including downloadable spreadsheets and PDFs, are converted
into a standardized CSV format using Optical Character Recognition (OCR)
tools such as Adobe Export PDF and ABBYY FineReader. Following
conversion, manual verification is performed to correct any OCR-induced
errors, ensuring the accuracy and consistency of the dataset.

The target population for this analysis includes all political donations
made to registered political parties, party leadership contestants,
riding associations (also known as electoral district or constituency
associations), and individual riding candidates during the specified
time frame. Donations are tracked across different types of electoral
events, including general elections, by-elections, and leadership races.
This comprehensive coverage allows for an in-depth examination of
donation patterns in various political contexts. The data collection
methodology ensures that all significant financial contributions are
captured, providing a robust foundation for analyzing the relationship
between political power and financial support.

@@

The data is organized into a structured format with each row detailing a
specific donation event, including the party receiving the donation, the
year it was made, the amount donated, and the constructed binary
indicators. This organization allows for efficient manipulation and
analysis using statistical programming tools like R. By aggregating
donations at the party-year level and incorporating relevant variables,
the dataset effectively translates real-world donation phenomena into
analyzable data points, enabling the investigation of how political
power status influences financial support in Canadian elections.

@@

The raw donation records underwent several preprocessing steps to ensure
the dataset's integrity and suitability for analysis. Initially, records
with missing or incomplete information in key variables such as Amount
Donated or Political Party were excluded from the analysis to maintain
data quality. This exclusion was essential to prevent potential biases
and inaccuracies in the subsequent analysis.

To ensure consistency across the dataset, political party names were
standardized. Variations like "Liberal Party of Ontario" and
"Progressive Conservative Party of Ontario" were consolidated under a
single standardized name, "Liberal Party of Ontario." This
standardization facilitated accurate aggregation and comparison across
different jurisdictions and time periods.

Donations were aggregated to reflect total amounts per party per year,
allowing for a higher-level analysis of donation patterns over time.
Binary variables such as Recipient_in_Power, Party_Size, and
Election_Year were created to streamline regression analyses, enabling
the examination of how political power status, party prominence, and
election cycles influence donation amounts.

Additional details on the data cleaning process, including specific
handling of anomalies, are provided in Appendix A: Data Cleaning Notes.

@@

The primary outcome variable in this analysis is the Amount Donated,
representing the monetary value of each donation made to a political
party within a given year. This continuous variable is essential for
assessing the level of financial support received by political parties
and understanding its relationship with factors such as power status and
party size.

The distribution of Amount Donated exhibits significant variability,
influenced by differing donor capacities and legal contribution limits
across jurisdictions. At the federal level, the average donation amount
is notably higher compared to the provincial level in Ontario,
reflecting the broader scope and potentially larger donor bases. This
difference underscores the importance of analyzing donation patterns
within the appropriate contextual framework.

Summary statistics for Amount Donated are presented in Table 1,
highlighting the total number of donations, the cumulative amount
donated, and the average donation per party per year.

>>>>

A key question in political finance is whether a party's position in power affects the amount of financial support it receives. Studies in other countries, such as the United States, have shown that incumbents often benefit from more donations, particularly from interest groups seeking to maintain influence (@fouirnaies2014financial). However, there has been less focus on whether this holds true in Canada, where the political system and donation regulations differ significantly. This paper aims to fill that gap by analyzing how a party's power status influences donor behaviour in federal and Ontario provincial elections.

@@

## Overview
The dataset used in this analysis comprises political donation records from 2013 to 2024, covering both the federal level and the province of Ontario in Canada. These records were sourced from the @ijf_political_donors Political Donors Dataset, which systematically collects and maintains comprehensive data on political donations across Canadian jurisdictions. The IJF updates this dataset daily by monitoring election agency websites at the federal level and within each province and territory, ensuring the data remains current and reliable.
The analysis was conducted using both R (@citeR) and Python (@python), leveraging a range of libraries for data manipulation and visualization. In R, key packages included dplyr (@dplyr), lubridate (@lubridate), tidyverse (@tidyverse), ggplot2 (@ggplot), readr (@readr), tibble (@tibble), and rstanarm (@rstanarm) for statistical modeling. In Python, pandas (@pandas), requests (@requests), and matplotlib (@matplotlib) were utilized for additional data processing and visualization tasks

@@

The dataset originates from @ijf_political_donors compilation of political donation records, which includes donations made to registered political parties, party leadership contestants, riding associations, and individual candidates. Donations are tracked across different types of electoral events, including general elections, by-elections, and leadership races. The IJF collects this data by monitoring official elections agency websites at both the federal level and within each province and territory.

The target population for this analysis includes all political donations made to registered political parties, party leadership contestants, riding associations (also known as electoral district or constituency associations), and individual riding candidates during the specified time frame. This comprehensive coverage allows for an in-depth examination of donation patterns in various political contexts. The data collection methodology ensures that all significant financial contributions are captured, providing a robust foundation for analyzing the relationship between political power and financial support.

@@

The data is organized into a structured format with each row detailing a specific donation event, including the party receiving the donation, the year it was made, the amount donated, and the constructed binary indicators. By aggregating donations at the party-year level and incorporating relevant variables, the dataset effectively translates real-world donation phenomena into analyzable data points, enabling the investigation of how political power status influences financial support in Canadian elections.

@@

The raw donation records were carefully preprocessed to ensure data integrity before analysis. First, the data was imported using the read_csv function from the readr package. Records with missing or incomplete information in key variables were excluded to maintain quality. Political party names were standardized to address inconsistencies, such as consolidating variations like "Liberal Party of Ontario" and "Ontario Liberal Party" under a single name. Donations were then aggregated at the party-year level to analyze total annual contributions received by each party. Binary variables, including Recipient_in_Power, Party_Size, and Election_Year, were created to support regression analysis. To normalize the distribution and handle zero values, a new variable, Log_Total_Donations, was generated by applying a logarithmic transformation to Total_Donations + 1. Finally, the cleaned dataset was cross-validated against the original records to ensure accuracy. Additional details about the data cleaning process can be found in Appendix \ref{sec-appendix}.

@@

The primary outcome variable is the Total Amount Donated, representing the total monetary value of donations received by each political party in a given year. This continuous variable is essential for assessing the level of financial support and understanding its relationship with factors such as power status and party size.

The distribution of total donation amounts varies between the federal and provincial levels. At the federal level, the average total donations are higher, reflecting broader donor bases and different legal contribution limits.
